\bf Accelerated and interpretable oblique random survival forests

### 2023-04-28 paper/jcgs/cover-letter/coverletter_main.tex
paper can be made using only public data

Journal of Biomedical Informatics

@@

We are writing to submit our manuscript ``\red{Accelerated and interpretable oblique random survival forests}'' to the \emph{Journal of Biomedical Informatics}.

@@

We suggest the following reviewers:
\begin{itemize}
    \item \red{Terry M. Therneau, Mayo Clinic (therneau@mayo.edu)}
    \item \red{Jack Dunn, Massachusetts Institute of Technology (jack.dunn.nz@gmail.com)}
    \item \red{Marvin N. Wright, University of Bremen (wright@leibniz-bips.de)}
\end{itemize}

@@

As the corresponding author, I confirm that none of the co-authors listed below have a conflict of interest with the action editors and referees I suggest above. Further, I confirm that all co-authors below consent to my submission of this manuscript to the \emph{Journal of Biomedical Informatics}.

>>>>

Journal of Computational and Graphical Statistics

@@

We are writing to submit our manuscript ``\red{Accelerated and interpretable oblique random survival forests}'' to the \emph{Journal of Computational and Graphical Statistics}.

@@

We suggest the following reviewers:
\begin{itemize}
    \item \red{Terry M. Therneau, Mayo Clinic (therneau@mayo.edu)}
    \item \red{Marvin N. Wright, University of Bremen (wright@leibniz-bips.de)}
\end{itemize}

@@

As the corresponding author, I confirm that none of the co-authors listed below have a conflict of interest with the action editors and referees I suggest above.

### 2023-04-28 paper/jcgs/main.Rnw
paper can be made using only public data

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

@@

A potential approach to improve the RSF when predictors are correlated and some are not relevant to the censored outcome is to use oblique trees instead of axis based trees. Axis based trees split data using a single predictor, creating decision boundaries that are perpendicular or parallel to axes of the predictor space \citep[see][Chapter~2]{breiman2017classification}. Oblique trees split data using a linear combination of predictors, creating decision boundaries that are neither parallel nor perpendicular to axes of their contributing predictors \citep[see][Chapter~5]{breiman2017classification}. \citet{menze2011oblique} examined prediction accuracy of RFs in the presence of correlated predictors and found that oblique RFs had substantially higher prediction accuracy compared to axis-based RFs. Similarly, \citet{jaeger2019oblique} found that growing RSFs with oblique rather than axis-based trees reduced the RSF's concordance error, with improvements ranging from 2.5\% to 24.9\% depending on the data analyzed.

@@

Building on prior research for bagging survival trees \citep{hothorn2004bagging}, \citet{hothorn2006unbiased} developed an axis-based RSF in their framework for unbiased recursive partitioning, more commonly referred to as the conditional inference forest (CIF). \citet{zhou2016random} developed a rotation forest based on the CIF and \citet{wang2017random} developed a method for extending the predictor space of the CIF. \citet{ishwaran2008random} developed an axis-based RSF with strict adherence to the rules for growing trees proposed in \citet{breiman2001random}.  \citet{jaeger2019oblique} developed the oblique RSF following the bootstrapping approach described in Breiman's original RF and incorporating early stopping rules from the CIF.

@@

Several techniques to estimate VI have been developed since \citet{breiman2001random} introduced permutation VI, which is defined for each predictor as the difference in a RF's estimated prediction error before versus after the predictor's values are randomly permuted. \citet{strobl2007bias} identified bias in permutation VI driven by variable selection bias and effects induced by bootstrap sampling, and proposed an unbiased permutation VI measure based on unbiased recursive partitioning \citep{hothorn2006unbiased}. \citet{menze2011oblique} introduced an approach to estimate VI for oblique RFs that computes an analysis of variance (ANOVA) table in non-leaf nodes to obtain p-values for each predictor contributing to the node. The ANOVA VI\footnote{\citet{menze2011oblique} name their method `oblique RF VI', but we use the name `ANOVA VI' in this article to avoid confusing Menze's approach with other approaches to estimate VI for oblique RFs.} is then defined for each predictor as the number of times a p-value associated with the predictor is $\leq 0.01$ while growing a forest. \citet{lundberg2017unified} introduced a method to estimate VI using SHapley Additive exPlanation (SHAP) values, which estimate the contribution of a predictor to a model's prediction for a given observation. SHAP VI is computed for each predictor by taking the mean absolute value of SHAP values for that predictor across all observations in a given set. With the exception of \citet{menze2011oblique}, few studies have evaluated estimation of VI using oblique RFs, and fewer have examined VI specifically for the oblique RSF.

@@

To demonstrate negation VI, consider a classification task where the goal is prediction of penguin species based on bill length and bill depth. Scaling these predictors to be centered at 0, we find oblique decision boundaries defined by linear combinations of bill length and bill depth correctly classify most of the data (Figure \ref{fig:negation_vs_permutation_viz}, Panel A). Permuting the values of bill length leads to several mis-classified observations, suggesting that bill length is an important predictor (Figure \ref{fig:negation_vs_permutation_viz}, Panel B). However, inspecting the permuted data shows that a number of observations moved to a region in the predictor space where there were previously no observations. Moving data to unobserved or perhaps unobservable regions of the predictor space may cause extrapolation error, which \citet{hooker2021unrestricted} identified as a cause of bias in permutation importance. Negating the coefficients for bill length in the linear combinations that define our decision boundaries causes the boundaries to flip across bill length's axis (Figure \ref{fig:negation_vs_permutation_viz}, Panel C). This leads to several mis-classified observations, suggesting that bill length is an important predictor without distorting the joint distribution of bill length and bill depth.

@@

<<negation_vs_permutation_viz, fig.height=4, fig.width=12, fig.cap="Demonstration of negation and permutation importance for a single oblique tree. Panel A shows the original data and decision boundaries. Panel B shows the data with permuted bill length values. Panel C shows the decision boundaries after negating coefficients for bill length. Permutation and negation both show that bill length is an important predictor, but permuting bill length distorts its joint distribution with bill depth.">>=

@@

p3 <- ggplot(data = penguins_permuted) +
  aes(x = x, y = y) +
  geom_point(aes(color = species, shape = species),
             size = 3,
             alpha = 0.8) +
  geom_polygon(data = poly1, fill = cols[1], color = 'black', alpha = 0.2) +
  geom_polygon(data = poly2, fill = cols[2], color = 'black', alpha = 0.2) +
  geom_polygon(data = poly3, fill = cols[3], color = 'black', alpha = 0.2) +
  geom_text(aes(x = -1.5, y = -1.5, label = "Impossible?"),
            size = 4, check_overlap = TRUE) +
  theme_minimal() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(x = "Bill length, scaled") +
  coord_cartesian(ylim = c(-2.1, 2.1),
                  xlim = c(-3.1, 3.1)) +
  theme(panel.grid = element_blank(),
        legend.position = '',
        text = element_text(size = 12),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())

@@

All analyses were conducted using R version \Sexpr{r_version} with version 0.0.4 of the \texttt{aorsf} \citep{jaeger2022aorsf} package. Analyses were coordinated with assistance from the \texttt{targets} package \citep{targets}. To standardize comparisons of computational efficiency, all learners and VI techniques used up to 4 processing units. All code used in the current study is available at \href{https://github.com/bcjaeger/aorsf-bench}{https://github.com/bcjaeger/aorsf-bench},

@@

<<bm_pred_viz_ibs, fig.height=10, fig.width=9, fig.cap="Index of prediction accuracy for the accelerated oblique random survival forest and other learning algorithms across multiple risk prediction tasks. Text appears in tasks where the accelerated oblique random survival forest obtained the highest index of prediction accuracy, showing the absolute and percent improvement over the second best learner. As predicted survival probabilities are not a standard output from \\texttt{xgboost-aft}, it is not included in this figure. Also, since this figure is intended to compare \\texttt{aorsf-fast} to learners that are not oblique random survival forests, \\texttt{aorsf-cph}, \\texttt{aorsf-net}, \\texttt{aorsf-random}, and \\texttt{obliqueRSF-net} are not included.">>=
bm_pred_viz$ibs_scaled$fig
@

<<bm_pred_model_viz_ibs, fig.height=10, fig.width=9, fig.cap="Expected differences in index of prediction accuracy between the accelerated oblique random survival forest and other learning algorithms. A region of practical equivalence is shown by purple dotted lines, and a boundary of non-zero difference is shown by an orange dotted line at the origin.">>=
bm_pred_model_viz$fig$ibs_scaled
@

@@

<<bm_pred_viz_cstat, fig.height=10, fig.width=9, fig.cap="Time-dependent concordance statistic for the accelerated oblique random survival forest and other learning algorithms across multiple risk prediction tasks. Text appears in tasks where the accelerated oblique random survival forest obtained the highest concordance, showing the absolute and percent improvement over the second best learner. Since this figure is intended to compare \\texttt{aorsf-fast} to learners that are not oblique random survival forests, \\texttt{aorsf-cph}, \\texttt{aorsf-net}, \\texttt{aorsf-random}, and \\texttt{obliqueRSF-net} are not included.">>=
bm_pred_viz$cstat$fig
@

<<bm_pred_model_viz_cstat, fig.height=10, fig.width=9, fig.cap="Expected differences in time-dependent concordance statistic between the accelerated oblique random survival forest and other learning algorithms. A region of practical equivalence is shown by purple dotted lines, and a boundary of non-zero difference is shown by an orange dotted line at the origin.">>=
bm_pred_model_viz$fig$cstat
@

@@

<<bm_vi_viz, fig.height=11, fig.width=9, fig.cap="Concordance statistic for assigning higher importance to relevant versus irrelevant variables. Text appears in rows where negation importance obtained the highest concordance, showing absolute and percent improvement over the second best technique.">>=
bm_vi_viz$fig
@

@@

In this paper, we have developed two contributions to the oblique RSF: (1) the accelerated oblique RSF (\ie, \texttt{aorsf-fast}) and (2) negation VI. Our technique to accelerate the oblique RSF reduces the number of operations required to find linear combinations of inputs using a single iteration of Newton Raphson scoring, while our VI technique directly engages with coefficients in linear combinations of inputs to measure importance of individual variables. In numeric experiments, we found that that \texttt{aorsf-fast} is approximately \Sexpr{times$obliqueRSF_net$ratio} times faster than its predecessor, \texttt{obliqueRSF-net}, with a practically equivalent C-statistic. We also found that negation VI, a technique to estimate VI using the oblique RSF, detected non-linear, combination, and main effects more effectively than three standard methods to estimate VI: permutation, ANOVA, and SHAP VI. Overall, we found that estimating VI using negation instead of ANOVA increased the C-statistic for ranking a relevant variable higher than an irrelevant variable by \Sexpr{cstat_aorsf_won_by$absolute}, a relative increase of \Sexpr{cstat_aorsf_won_by$percent}\%.

>>>>

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Accelerated and interpretable oblique random survival forests}
\end{center}
  \medskip
} \fi

@@

A potential approach to improve the RSF when predictors are correlated and some are not relevant to the censored outcome is to use oblique trees instead of axis based trees. Axis based trees split data using a single predictor, creating decision boundaries that are perpendicular or parallel to axes of the predictor space \citep[see][Chapter~2]{breiman2017classification}. Oblique trees split data using a linear combination of predictors, creating decision boundaries that are neither parallel nor perpendicular to axes of their contributing predictors \citep[see][Chapter~5]{breiman2017classification}. Oblique trees may create more adequate partitions of a predictor space compared to axis-based trees, as demonstrated in Figure \ref{fig:axis_v_oblique}. \citet{menze2011oblique} examined prediction accuracy of RFs in the presence of correlated predictors and found that oblique RFs had substantially higher prediction accuracy compared to axis-based RFs. Similarly, \citet{jaeger2019oblique} found that growing RSFs with oblique rather than axis-based trees reduced the RSF's concordance error, with improvements ranging from 2.5\% to 24.9\% depending on the data analyzed.

<<axis_v_oblique, fig.height=5, fig.width=12, fig.cap="Decision boundaries from an axis based (panel A) and oblique (panel B) decision tree used to classify penguin species based on bill depth and bill length. The decision boundary from the oblique tree is better able to capture the geometry of this data, leading to fewer mis-classified penguins.">>=

poly1_axis <- data.frame(
 x = c(-3, 0.2116068, 0.2116068, -3, -3),
 y = c(-2, -2,   2, 2, -2)
)

poly2_axis <- data.frame(
 x = c(0.2116068, 3, 3, 0.2116068, 0.2116068),
 y = c(-0.4137976, -0.4137976, -2, -2, -0.4137976)
)

poly3_axis <- data.frame(
 x = c(0.2116068, 3, 3, 0.2116068, 0.2116068),
 y = c(2, 2, -0.4137976, -0.4137976, 2)
)

p1 <- ggplot(data = penguins_scaled) +
  aes(x = x, y = y) +
  geom_point(aes(color = species, shape = species),
             size = 3,
             alpha = 0.8) +
  geom_polygon(data=poly1_axis, fill=cols[1], color='black', alpha=0.2) +
  geom_polygon(data=poly2_axis, fill=cols[2], color='black', alpha=0.2) +
  geom_polygon(data=poly3_axis, fill=cols[3], color='black', alpha=0.2) +
  theme_minimal() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(y = "Bill depth, scaled",
       x = "Bill length, scaled") +
  coord_cartesian(ylim = c(-2.1, 2.1),
                  xlim = c(-3.1, 3.1)) +
  theme(panel.grid = element_blank(),
        legend.position = '',
        text = element_text(size = 12))

p2 <- ggplot(data = penguins_scaled) +
  aes(x = x, y = y) +
  geom_point(aes(color = species, shape = species),
             size = 3,
             alpha = 0.8) +
  geom_polygon(data=poly1, fill=cols[1], color='black', alpha=0.2) +
  geom_polygon(data=poly2, fill=cols[2], color='black', alpha=0.2) +
  geom_polygon(data=poly3, fill=cols[3], color='black', alpha=0.2) +
  theme_minimal() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(y = "",
       x = "Bill length, scaled") +
  coord_cartesian(ylim = c(-2.1, 2.1),
                  xlim = c(-3.1, 3.1)) +
  theme(panel.grid = element_blank(),
        legend.position = '',
        text = element_text(size = 12))

p3 <- ggarrange(tag_facet(p1, tag_pool = "A", vjust = 2.5),
                tag_facet(p2, tag_pool = "B", vjust = 2.5),
                nrow = 1)

p3

@@

\citet{hothorn2006unbiased} developed an axis-based RSF in their framework for unbiased recursive partitioning, more commonly referred to as the conditional inference forest (CIF). \citet{zhou2016random} developed a rotation forest based on the CIF and \citet{wang2017random} developed a method for extending the predictor space of the CIF. \citet{ishwaran2008random} developed an axis-based RSF with strict adherence to the rules for growing trees proposed in \citet{breiman2001random}.  \citet{jaeger2019oblique} developed the oblique RSF following the bootstrapping approach described in Breiman's original RF and incorporating early stopping rules from the CIF.

@@

Several techniques to estimate VI have been developed since \citet{breiman2001random} introduced permutation VI, which is defined for each predictor as the difference in a RF's estimated prediction error before versus after the predictor's values are randomly permuted. \citet{strobl2007bias} identified bias in permutation VI driven by categorical predictors and bootstrap sampling, and proposed a permutation VI measure based on unbiased recursive partitioning \citep{hothorn2006unbiased}. \citet{menze2011oblique} introduced an approach to estimate VI for oblique RFs that computes an analysis of variance (ANOVA) table in non-leaf nodes to obtain p-values for each predictor contributing to the node. The ANOVA VI\footnote{\citet{menze2011oblique} name their method `oblique RF VI', but we use the name `ANOVA VI' in this article to avoid confusing Menze's approach with other approaches to estimate VI for oblique RFs.} is then defined for each predictor as the number of times a p-value associated with the predictor is $\leq 0.01$ while growing a forest. \citet{lundberg2017unified} introduced a method to estimate VI using SHapley Additive exPlanation (SHAP) values, which estimates the contribution of a predictor to a model's prediction for a given observation. SHAP VI is computed for each predictor by taking the mean absolute value of SHAP values for that predictor across all observations in a given set. With the exception of \citet{menze2011oblique}, few studies have evaluated estimation of VI using oblique RFs, and fewer have examined VI specifically for the oblique RSF.

@@

To demonstrate negation VI, consider a classification task where the goal is prediction of penguin species (chinstrap, gentoo, or adelie) based on bill length and bill depth \citep{palmerpenguins_rpack}. Scaling these predictors to be centered at 0, we find oblique decision boundaries defined by linear combinations of bill length and bill depth correctly classify most of the data (Figure \ref{fig:negation}, Panel A). Permuting the values of bill length leads to several mis-classified observations, suggesting that bill length is an important predictor (Figure \ref{fig:negation}, Panel B). However, inspecting the permuted data shows that a number of observations moved to a region in the predictor space where there were previously no observations. Moving data to unobserved or perhaps unobservable regions of the predictor space may cause extrapolation error, which \citet{hooker2021unrestricted} identified as a cause of bias in permutation importance. Negating the coefficients for bill length in the linear combinations that define our decision boundaries causes the boundaries to flip across bill length's axis (Figure \ref{fig:negation}, Panel C). This leads to several mis-classified observations, suggesting that bill length is an important predictor without distorting the joint distribution of bill length and bill depth.

@@

<<negation, fig.height=4, fig.width=12, fig.cap="Demonstration of negation and permutation importance for a single oblique tree. Panel A shows the original data and decision boundaries. Panel B shows the data with permuted bill length values. Panel C shows the decision boundaries after negating coefficients for bill length. Permutation and negation both show that bill length is an important predictor, but permuting bill length distorts its joint distribution with bill depth.">>=

@@

p3 <- ggplot(data = penguins_permuted) +
  aes(x = x, y = y) +
  geom_point(aes(color = species, shape = species),
             size = 3,
             alpha = 0.8) +
  geom_polygon(data = poly1, fill = cols[1], color = 'black', alpha = 0.2) +
  geom_polygon(data = poly2, fill = cols[2], color = 'black', alpha = 0.2) +
  geom_polygon(data = poly3, fill = cols[3], color = 'black', alpha = 0.2) +
  # geom_text(aes(x = -2, y = -1.5, label = "Impossible?"),
            # size = 4, check_overlap = TRUE) +
  theme_minimal() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(x = "Bill length, scaled") +
  coord_cartesian(ylim = c(-2.1, 2.1),
                  xlim = c(-3.1, 3.1)) +
  theme(panel.grid = element_blank(),
        legend.position = '',
        text = element_text(size = 12),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())

@@

All analyses were conducted using R version \Sexpr{r_version} with version 0.0.4 of the \texttt{aorsf} \citep{jaeger2022aorsf} package. Analyses were coordinated with assistance from the \texttt{targets} package \citep{targets}. To standardize comparisons of computational efficiency, all learners and VI techniques used up to 4 processing units.

% All code used in the current study is available at \href{https://github.com/bcjaeger/aorsf-bench}{https://github.com/bcjaeger/aorsf-bench},

@@

<<bm_pred_viz_ibs, fig.height=9, fig.width=9, fig.cap="Index of prediction accuracy in multiple risk prediction tasks. Text appears in tasks where the accelerated oblique random survival forest obtained the highest score, showing absolute and relative improvement over the second best learner. Since this figure is intended to compare \\texttt{aorsf-fast} to learners that are not oblique random survival forests, \\texttt{aorsf-cph}, \\texttt{aorsf-random}, and \\texttt{obliqueRSF-net} are not included.">>=
bm_pred_viz$ibs_scaled$fig
@

<<bm_pred_model_viz_ibs, fig.height=9, fig.width=9, fig.cap="Expected differences in index of prediction accuracy between the accelerated oblique random survival forest and other learning algorithms. A region of practical equivalence is shown by purple dotted lines, and a boundary of non-zero difference is shown by an orange dotted line at the origin.">>=
bm_pred_model_viz$fig$ibs_scaled
@

@@

<<bm_pred_viz_cstat, fig.height=9, fig.width=9, fig.cap="Time-dependent concordance statistic for the accelerated oblique random survival forest and other learning algorithms across multiple risk prediction tasks. Text appears in tasks where the accelerated oblique random survival forest obtained the highest concordance, showing the absolute and percent improvement over the second best learner. Since this figure is intended to compare \\texttt{aorsf-fast} to learners that are not oblique random survival forests, \\texttt{aorsf-cph}, \\texttt{aorsf-random}, and \\texttt{obliqueRSF-net} are not included.">>=
bm_pred_viz$cstat$fig
@

<<bm_pred_model_viz_cstat, fig.height=9, fig.width=9, fig.cap="Expected differences in time-dependent concordance statistic between the accelerated oblique random survival forest and other learning algorithms. A region of practical equivalence is shown by purple dotted lines, and a boundary of non-zero difference is shown by an orange dotted line at the origin.">>=
bm_pred_model_viz$fig$cstat
@

@@

<<bm_vi_viz, fig.height=10, fig.width=9, fig.cap="Concordance statistic for assigning higher importance to relevant versus irrelevant variables. Text appears in rows where negation importance obtained the highest concordance, showing absolute and percent improvement over the second best technique.">>=
bm_vi_viz$fig
@

@@

In this paper, we have developed two contributions to the oblique RSF: (1) the accelerated oblique RSF (\ie, \texttt{aorsf-fast}) and (2) negation VI. Our technique to accelerate the oblique RSF reduces the number of operations required to find linear combinations of inputs using a single iteration of Newton Raphson scoring, while our VI technique directly engages with coefficients in linear combinations of inputs to measure importance of individual variables. In numeric experiments, we found that \texttt{aorsf-fast} is approximately \Sexpr{times$obliqueRSF_net$ratio} times faster than its predecessor, \texttt{obliqueRSF-net}, with a practically equivalent C-statistic. We also found that negation VI, a technique to estimate VI using the oblique RSF, detected non-linear, combination, and main effects more effectively than three standard methods to estimate VI: permutation, ANOVA, and SHAP VI. Overall, we found that estimating VI using negation instead of ANOVA increased the C-statistic for ranking a relevant variable higher than an irrelevant variable by \Sexpr{cstat_aorsf_won_by$absolute}, a relative increase of \Sexpr{cstat_aorsf_won_by$percent}\%.
